\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

\title{GAD+ Experiment Report:\\Transition State Search with Eckart-Projected GAD and HIP}
\author{GAD\_plus Project}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
We benchmark Gentlest Ascent Dynamics (GAD) for transition state (TS) search
using the HIP neural network potential on the Transition1x dataset (9,561 organic
reactions). Eckart-projected GAD (Level~2) with $\mathrm{dt}=0.01$ achieves 62\%
convergence at noise levels from 10--100\,pm, compared to 0\% for unprojected GAD
at $\geq50$\,pm. The basin of attraction extends to $\sim100$\,pm: when GAD converges,
it reliably returns to the correct TS. We test seven method variants including
NR-GAD ping-pong, adaptive timestep, and displacement clamping to push convergence
beyond the 62\% plateau.
\end{abstract}

\tableofcontents

%----------------------------------------------------------------------
\section{Background}
\label{sec:background}

\subsection{Gentlest Ascent Dynamics (GAD)}

GAD modifies atomic forces to ascend along the lowest Hessian eigenvector while
descending along all others. At each step:
\begin{enumerate}
    \item Compute energy $E$, forces $\mathbf{F} = -\nabla E$, and full Hessian $\mathbf{H}$
          via the HIP neural network potential.
    \item Eckart-project $\mathbf{H}$ to remove 6 translational/rotational modes
          (3 translation + 3 rotation for nonlinear molecules), giving vibrational
          eigenvalues $\{\lambda_i\}$ and eigenvectors $\{\mathbf{v}_i\}$.
    \item Identify the lowest vibrational eigenvector $\mathbf{v}_0$ as the ascent direction.
    \item Apply the GAD formula:
          $\mathbf{F}_{\mathrm{GAD}} = \mathbf{F} + 2(\mathbf{F} \cdot \mathbf{v}_0)\mathbf{v}_0$
    \item Take an Euler step: $\mathbf{x}_{n+1} = \mathbf{x}_n + \mathrm{dt} \cdot \mathbf{F}_{\mathrm{GAD}}$
\end{enumerate}

\subsection{Convergence Criterion}

A geometry is a converged transition state when:
\begin{equation}
    n_{\mathrm{neg}} = 1 \quad \text{AND} \quad \|\mathbf{F}\| < 0.01 \;\text{eV/\AA}
\end{equation}
where $n_{\mathrm{neg}}$ is the number of negative eigenvalues in the Eckart-projected
vibrational Hessian (Morse index~1). This is non-negotiable---no eigenvalue product
gates, no threshold relaxation.

\subsection{Feature Levels}

\begin{table}[h]
\centering
\caption{GAD feature levels. Each level adds one feature to the base.}
\begin{tabular}{clp{8cm}}
\toprule
Level & Name & Description \\
\midrule
0 & \texttt{pure\_gad} & Raw Hessian, fixed $\mathrm{dt}=0.005$, no mode tracking \\
1 & \texttt{gad\_tracked} & + Mode tracking ($k=8$) \\
2 & \texttt{gad\_projected} & + Eckart projection in GAD dynamics \\
3 & \texttt{gad\_adaptive\_dt} & + Eigenvalue-clamped adaptive $\mathrm{dt}$ \\
4 & \texttt{nr\_gad\_flipflop} & + Newton-Raphson refinement when $n_{\mathrm{neg}}=1$ \\
\bottomrule
\end{tabular}
\label{tab:levels}
\end{table}

\subsection{Experimental Setup}

\begin{itemize}
    \item \textbf{Potential:} HIP neural network (\texttt{hip\_v2.ckpt}), direct analytical Hessian
    \item \textbf{Dataset:} Transition1x train split (9,561 samples, organic reactions)
    \item \textbf{Cluster:} Narval (Alliance Canada), A100 MIG slices (10\,GB VRAM each)
    \item \textbf{Timing:} $\sim0.06$\,s per GAD step; 300~steps $\approx 18$\,s per molecule
    \item \textbf{Starting geometry:} Known TS + Gaussian noise (unless otherwise noted)
    \item \textbf{All methods are geometry-based} (state-based, no path history)
\end{itemize}

%----------------------------------------------------------------------
\section{Phase 1: Parameter Sweep}
\label{sec:phase1}

\textbf{Question:} What timestep and mode tracking window work best for Level~2?

\textbf{Setup:} 10 test-split samples, 100 steps, 50\,pm noise.
Grid: $\mathrm{dt} \in \{0.001, 0.003, 0.005, 0.01, 0.02\}$,
$k_{\mathrm{track}} \in \{0, 4, 8\}$.

\begin{table}[h]
\centering
\caption{Phase~1: Convergence rate by $\mathrm{dt}$ and $k_{\mathrm{track}}$.
Only $\mathrm{dt}=0.01$ achieves any convergence at 50\,pm noise in 100 steps.}
\begin{tabular}{ccccc}
\toprule
$\mathrm{dt}$ & $k_{\mathrm{track}}=0$ & $k_{\mathrm{track}}=4$ & $k_{\mathrm{track}}=8$ & Best \\
\midrule
0.001 & 0/10 & 0/10 & 0/10 & --- \\
0.003 & 0/10 & 0/10 & 0/10 & --- \\
0.005 & 0/10 & 0/10 & 0/10 & --- \\
\textbf{0.01} & \textbf{4/10 (40\%)} & 3/10 & 3/10 & $\checkmark$ \\
0.02  & 0/10 & 0/10 & 0/10 & --- \\
\bottomrule
\end{tabular}
\label{tab:phase1}
\end{table}

\textbf{Finding:} $\mathrm{dt}=0.01$ is optimal (2$\times$ the default). Mode tracking
($k_{\mathrm{track}}$) has negligible effect when Eckart projection is active---the
projection already handles eigenvector consistency.

\textbf{Data:} \texttt{/lustre07/scratch/memoozd/gadplus/runs/sweep\_dt/}

%----------------------------------------------------------------------
\section{Phase 2: Noise Robustness (300 samples)}
\label{sec:phase2}

\textbf{Question:} How does convergence degrade with increasing noise on the TS geometry?

\textbf{Setup:} 300 train-split samples, 300 steps, $\mathrm{dt}=0.01$, Level~2.
9~noise levels from 0--200\,pm, run as 9 parallel MIG jobs.

\begin{table}[h]
\centering
\caption{Phase~2: Convergence rate vs.\ noise level (300 samples).
\textcolor{red}{TO BE UPDATED with 300-sample results.}
Preliminary 50-sample results shown.}
\begin{tabular}{rrrrl}
\toprule
Noise (pm) & Conv & Rate & Avg Steps & Notes \\
\midrule
0   & 48/50 & 96\%  & 9   & Near-perfect at known TS \\
10  & 32/50 & 64\%  & 31  & First significant drop \\
20  & 31/50 & 62\%  & 45  & Plateau begins \\
30  & 31/50 & 62\%  & 65  & \\
50  & 31/50 & 62\%  & 85  & Level~0 gets 0\% here \\
70  & 31/50 & 62\%  & 94  & Stable \\
100 & 31/50 & 62\%  & 106 & Still on plateau \\
150 & 23/50 & 46\%  & 119 & Degradation begins \\
200 & 17/50 & 34\%  & 142 & \\
\bottomrule
\end{tabular}
\label{tab:phase2}
\end{table}

\textbf{Key finding:} Level~2 maintains $\sim62\%$ convergence across 10--100\,pm.
Level~0 drops to 0\% by 50\,pm. The Eckart projection is the single biggest
algorithmic improvement. The 38\% failure plateau suggests structural barriers
that projection alone cannot overcome.

\textbf{Data:} \texttt{/lustre07/scratch/memoozd/gadplus/runs/noise\_survey\_300/}

%----------------------------------------------------------------------
\section{Phase 3: Starting Geometry (300 samples)}
\label{sec:phase3}

\textbf{Question:} Can GAD find a TS from geometries other than a noised TS?

\textbf{Setup:} 300 train-split samples, 300 steps, $\mathrm{dt}=0.01$, Level~2.

\begin{table}[h]
\centering
\caption{Phase~3: Convergence by starting geometry (300 samples).
\textcolor{red}{TO BE UPDATED with 300-sample results.}
Preliminary 50-sample results shown.}
\begin{tabular}{lrrrr}
\toprule
Starting Geometry & Conv & Rate & Avg Steps & Avg $n_{\mathrm{neg}}$ \\
\midrule
Noised TS (10\,pm) & 32/50 & 64\% & 31  & 1.3 \\
Midpoint R$\to$P   & 15/50 & 30\% & 174 & 1.5 \\
Reactant           & 7/50  & 14\% & 281 & 0.7 \\
Product            & 2/50  & 4\%  & 141 & 1.0 \\
\bottomrule
\end{tabular}
\label{tab:phase3}
\end{table}

\textbf{Finding:} Reactant starts have $n_{\mathrm{neg}} \approx 0$ (stuck at minimum).
GAD fundamentally requires starting near a saddle region. Midpoint interpolation
is a viable alternative (30\%) because it often lands in a high-energy region
near the TS.

\textbf{Data:} \texttt{/lustre07/scratch/memoozd/gadplus/runs/starting\_geom\_300/}

%----------------------------------------------------------------------
\section{Phase 5: IRC Validation (10 samples, preliminary)}
\label{sec:phase5}

\textbf{Question:} When GAD converges, is it the \emph{intended} TS?

From each converged TS, run Sella IRC forward/backward. Compare endpoints to
known reactant/product via RMSD (threshold 0.3\,\AA).

\begin{table}[h]
\centering
\caption{Phase~5: IRC validation at noise $= 10$\,pm (10 samples).
Larger-scale validation (30 samples $\times$ 3 noise levels) pending.}
\begin{tabular}{lr}
\toprule
Classification & Count \\
\midrule
Intended (both matched)    & 3/10 \\
Half-intended (one matched) & 4/10 \\
Unintended (neither matched) & 3/10 \\
\bottomrule
\end{tabular}
\label{tab:phase5}
\end{table}

\textbf{Finding:} Half-intended cases consistently match the product (RMSD$\sim$0.16\,\AA)
but not the reactant (RMSD$\sim$0.45\,\AA). The TS is real but may connect a
different reactant conformer. RMSD threshold 0.3\,\AA\ may be too tight for
reactions with large conformational changes.

%----------------------------------------------------------------------
\section{Phase 6: Basin Mapping (50 samples)}
\label{sec:phase6}

\textbf{Question:} At what noise level do we start finding \emph{different} TS?

Start from known TS, add increasing noise, check if converged TS matches original
(RMSD $< 0.1$\,\AA).

\begin{table}[h]
\centering
\caption{Phase~6: Basin stability (50 samples).
\textcolor{red}{TO BE UPDATED with 50-sample results.}
Preliminary 20-sample results shown.}
\begin{tabular}{rrrrr}
\toprule
Noise (pm) & Conv & Same TS & Diff TS & Avg RMSD (\AA) \\
\midrule
0   & 19/20 & 19 & 0 & 0.0005 \\
10  & 13/20 & 13 & 0 & 0.0060 \\
20  & 14/20 & 14 & 0 & 0.0105 \\
50  & 13/20 & 13 & 0 & 0.0243 \\
100 & 12/20 & 12 & 0 & 0.0452 \\
200 & 2/20  & 1  & 1 & 0.1178 \\
500 & 0/20  & 0  & 0 & --- \\
\bottomrule
\end{tabular}
\label{tab:phase6}
\end{table}

\textbf{Key finding:} Basin width $\sim100$\,pm. When GAD converges, it always returns
to the same TS up to 100\,pm. First different TS at 200\,pm. This means the 62\%
failure plateau is genuine non-convergence, not silent wrong answers.

%----------------------------------------------------------------------
\section{Phase 7: Method Comparison (7 variants)}
\label{sec:phase7}

\textbf{Question:} Can we push past the 62\% plateau with better GAD variants?

\textbf{Setup:} 50 train-split samples, 300 steps, 6 noise levels
($10, 30, 50, 100, 150, 200$\,pm), 7 methods:

\begin{enumerate}
    \item \textbf{gad\_projected} --- Level~2 baseline ($\mathrm{dt}=0.01$, fixed)
    \item \textbf{gad\_adaptive\_dt} --- Eigenvalue-clamped: $\mathrm{dt} \propto 1/\mathrm{clamp}(|\lambda_0|)$
    \item \textbf{gad\_tight\_clamp} --- Max per-atom displacement capped at 0.1\,\AA\ (vs.\ 0.35\,\AA\ default)
    \item \textbf{gad\_adaptive\_tight} --- Adaptive $\mathrm{dt}$ + tight clamp combined
    \item \textbf{gad\_small\_dt} --- Conservative $\mathrm{dt}=0.005$ for stability comparison
    \item \textbf{nr\_gad\_pingpong} --- NR minimize when $n_{\mathrm{neg}} \geq 2$,
          GAD when $n_{\mathrm{neg}} < 2$
    \item \textbf{nr\_gad\_pp\_adaptive} --- Ping-pong + adaptive $\mathrm{dt}$ in GAD phase
\end{enumerate}

The NR step is pure Newton descent: $\Delta\mathbf{x} = -\mathbf{H}^{-1}\mathbf{g}$
in the vibrational subspace, with eigenvalue flooring for regularization. No RFO,
no trust radius. All methods are geometry-based (no path history).

\begin{table}[h]
\centering
\caption{Phase~7: Method comparison across noise levels (50 samples each).
\textcolor{red}{RESULTS PENDING --- jobs submitted as SLURM array 58835900.}}
\begin{tabular}{lcccccc}
\toprule
Method & 10\,pm & 30\,pm & 50\,pm & 100\,pm & 150\,pm & 200\,pm \\
\midrule
gad\_projected       & --- & --- & --- & --- & --- & --- \\
gad\_adaptive\_dt    & --- & --- & --- & --- & --- & --- \\
gad\_tight\_clamp    & --- & --- & --- & --- & --- & --- \\
gad\_adaptive\_tight & --- & --- & --- & --- & --- & --- \\
gad\_small\_dt       & --- & --- & --- & --- & --- & --- \\
nr\_gad\_pingpong    & --- & --- & --- & --- & --- & --- \\
nr\_gad\_pp\_adaptive & --- & --- & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\label{tab:phase7}
\end{table}

\textbf{Hypotheses:}
\begin{itemize}
    \item NR-GAD ping-pong should help at high noise ($\geq100$\,pm) where many samples
          start at higher-order saddles ($n_{\mathrm{neg}} \geq 2$). At low noise it
          should match the baseline (almost always $n_{\mathrm{neg}} < 2$).
    \item Adaptive $\mathrm{dt}$ should help in steep-curvature regions by automatically
          reducing the step size, preventing overshooting.
    \item Tight clamping should reduce oscillatory failures but may slow convergence.
\end{itemize}

%----------------------------------------------------------------------
\section{Summary of Findings}
\label{sec:summary}

\begin{enumerate}
    \item \textbf{Eckart projection is the biggest win.} Level~2 vs Level~0 at 50\,pm:
          62\% vs 0\%. The projection correctly removes TR modes and gives physically
          meaningful $n_{\mathrm{neg}}$ counts.

    \item \textbf{$\mathrm{dt}=0.01$ is optimal} for Level~2 (2$\times$ the default).
          Mode tracking ($k_{\mathrm{track}}$) is redundant when projection is active.

    \item \textbf{62\% convergence plateau} from 10--100\,pm. Steps scale linearly with
          noise (9 steps at 0\,pm $\to$ 142 at 200\,pm).

    \item \textbf{Basin of attraction $\sim100$\,pm.} GAD either converges to the correct
          TS or fails to converge at all. It does not silently find wrong answers.

    \item \textbf{Starting from minima rarely works.} Reactant start: 14\%, product: 4\%.
          GAD needs to start in a saddle region.

    \item \textbf{IRC validation:} 30\% fully intended, 40\% half-intended.
          Half-intended cases find a real TS but from a slightly different pathway.
\end{enumerate}

%----------------------------------------------------------------------
\section{Reproducing Results}
\label{sec:reproduce}

All scripts are standalone (argparse, no Hydra) in \texttt{scripts/}.
Submit via \texttt{sbatch}:

\begin{verbatim}
# Phase 1: Parameter sweep (1 job, ~20 min)
sbatch scripts/run_sweep_dt.slurm

# Phase 2: Noise survey, 300 samples (9 parallel MIG jobs, ~90 min)
sbatch scripts/run_noise_survey.slurm

# Phase 3: Starting geometry, 300 samples (4 parallel jobs, ~90 min)
sbatch scripts/run_starting_geom.slurm

# Phase 6: Basin mapping, 50 samples (1 job, ~2 hrs)
sbatch scripts/run_basin_map.slurm

# Phase 7: Method comparison (6 parallel jobs, ~2 hrs)
sbatch scripts/run_method_comparison.slurm
\end{verbatim}

Results are Parquet files queryable via DuckDB:
\begin{verbatim}
SELECT noise_pm, COUNT(*) as total,
       SUM(CASE WHEN converged THEN 1 ELSE 0 END) as conv,
       ROUND(100.0 * conv / total, 1) as rate
FROM '/lustre07/scratch/.../summary_*.parquet'
GROUP BY noise_pm ORDER BY noise_pm;
\end{verbatim}

\end{document}
